---
title: "R Notebook"
output: html_notebook
---

The following MATLAB code implements the `Pettis` algorithm for estimating
intrinsic dimensionality, which is also contained in the EDA Toolbox function called `idpettis`. We first generate some data to illustrate the functionality of the algorithm. The helix is described by the following equations, and points are randomly chosen along this path:

$$
x = cos(\theta) \\
y = sin(\theta) \\
z = 0.1 \theta
$$
for $0<= \theta <= 4 \pi$.

For this data set, the dimensionality is 3, but the intrinsic dimensionality is 1. We show a picture of the helix in Figure 2.6. We obtain the data by generating uniform random numbers in the interval￿$0<= \theta <= 4 \pi$.

### Set up project folders with `rprojroot` package

```{r results='hold'}
# package rprojroot required
library(rprojroot)

root <- rprojroot::is_rstudio_project
root_file <- root$make_fix_file()
# root$find_file("REAME.md")
root_file()
cat("Setting up the project folders:\n")
project.data <- root_file("data")
project.extdata <- root_file("inst/extdata")
project.R <- root_file("R")
project.matlab <- root_file("inst/matlab")

project.data
project.extdata
project.R
project.matlab
```

```{r fig.asp=1}
library("plot3D")

# generate the random numbers
# use random numbers from the uniform distribution
n <- 500
theta <- runif(n =  n, min = 0, max = 4*pi)


# Use in the equations for a helix￿
x <- cos(theta)
y <- sin(theta)
z <- 0.1 * (theta)

# Put into a data matrix
# X = [x(,),y(,),z(,)]
X <- matrix(c(x, y, z), 500, 3)

# plot it
scatter3D(x, y, z,
          col = "blue",
          pch = 19,
          cex = 0.5,
          bty = "b2",
          xlab = "x",
          ylab = "y",
          zlab = "z",
          theta = -45,       # 3D viewing direction: azimuth
          phi = 30,           # co-latitude
          ticktype = "detailed"
          )
```

